Computing Direct Sum Decompositions

### 2025-04-23 paper/main.tex
finished the remark

\begin{lem}\label{lem:graded_algorithm}
  Let $N = \mu(M)$ and set $\psi = \phi^N$. Then $M = \im\psi \oplus \coker\psi$.
\end{lem}
\begin{proof}
  Fix homogeneous generators $m_1,\dots,m_N$ of $M$, with $\deg m_i$ nonincreasing, and let $d_1,\dots,d_n$ be the decreasing list of unique degrees of generators of $M$. The matrix for $\phi$ is then ``block upper triangular'' with $n$ diagonal blocks with entries in $k$, i.e., of the form
  \[ \phi = \begin{pmatrix}
    B_1    & *      & \dots  & *      \\
    0      & B_2    & \dots  & *      \\
    \vdots & \vdots & \ddots & \vdots \\
    0      & 0      & \dots  & B_n
  \end{pmatrix}, \]
  where the blocks $B_i$ are square matrices \emph{with entries in $k$}, and the remaining entries are in $R$. To see this, note that since $\phi$ has degree zero, $\phi(m_i)$ is a $k$-linear combination of the $m_j$ with $\deg m_j = \deg m_i$ plus an $R$-linear combination of the $m_j$ with $\deg m_j < \deg m_i$. In particular, $B_i$ is the induced endomorphism on the submodule generated by all generators of degree $d_i$.

@@

  Let $\alpha$ be the composition of these automorphisms for each block $B_i$, then $\phi' = \alpha^{-1}\circ\phi\circ\alpha$ corresponds to performing row and column operations on $\phi$ over $k$ and has the form
  \begin{equation}
    \label{eq:Jordan}
    \phi' = \begin{pmatrix}
      J_1    & *      & \dots  & *      \\
      0      & J_2    & \dots  & *      \\
      \vdots & \vdots & \ddots & \vdots \\
      0      & 0      & \dots  & J_n
    \end{pmatrix},
  \end{equation}
  where each $J_i$ is a Jordan block over $k$ with eigenvalue $\lambda_i$, and the remaining entries are in $R$.
  Relabeling, we may assume that $\lambda_i\neq 0$ for $i=1,\dots,n_0$ and $\lambda_i=0$ for $i=n_0+1,\dots,n$.

  If we take $N_0$ to be at least the sum of the sizes of the nilpotent Jordan blocks $J_{n_0+1},\dots,J_n$, then for $N\geq N_0$ we have
  $$
  \phi^N =
  \begin{pmatrix}
    J_1^N & *** &\dots & \dots& \dots  &\dots & *** \\
    0 & J_2^N & \dots &\dots&\dots & \dots& *** \\
    \vdots &\vdots & \ddots &\vdots \\
    0 & 0 & \dots & J_{n_0}^N & *** & \dots & ***\\
    0 & 0 & \dots & 0 & 0 & \dots & 0\\
    \vdots &\vdots & \ddots &\vdots \\
    0 & 0 & \dots & 0 & 0 & \dots & 0
  \end{pmatrix},
  $$
  where each nonzero $J_i^N$ has the units $\lambda_i^N$ on the diagonal.
  ($\mu(M)$ is an upper bound on the sum of the sizes of the nilpotent Jordan blocks, so the above holds for $N\geq \mu(M)$.)

  Fix $N$ as above.
  Now, consider the map $ \phi^N: \im(\phi^N) \to \im(\phi^N)$. If this map is surjective, then $\im(\phi^N)$ is a direct summand of $M$, with complementary summand $\coker \phi^N$: to see this, observe that the injection $\im(\phi^N) \inc M$ would then  compose with $\phi^N$ to be a surjection, and since $\im(\phi^N)$ is a finitely generated module, a surjection must be an isomorphism $\a:\im \phi^N \to \im \phi^N$ by \cite[Corollary~4.4]{Eisenbud95}. Thus, we obtain a splitting of the inclusion $\im \phi^N \inc M$ via $\a\inv\circ \phi^N$; since $\a\inv$ is an isomorphism, we have that $\im \phi^N$ is (isomorphic to) a direct summand of $M$, with complementary summand $\coker \phi^N$.

  So, all that needs to be seen is that $\psi:=\psi^N$ is surjective.
  To see this, note that after replacing the original generators for $M$ by the generators that put $\psi$ into the form above,  a generating set for $\im(\phi^N)$ is given by $f_i = \psi(m_i) = u_i\cdot e_i + \sum_{j<i } a_{ij} e_j$ for $i = 1,\dots,n-N_0$, where $u_i$ is a unit (explicitly, $u_i$ is a power of some nonzero $\lambda_{i'}\in k$).
  That is, a generating set is given by
  $$u_1 e_1,\quad  u_2 e_2+a_{21}e_1,\quad \dots ,\quad u_{n-n_0}e_{n-N_0}+a_{n-N_0,n-1}e_{n-N_0-1}+\dots + a_{N-n_0,1}e_1.$$
  Since the $u_i$ are units, the $R$-span of these generators is clearly the same as the $R$-span of $e_1,\dots,e_{n-N_0}$, and thus after applying $\psi$ we get back the $R$-span of $f_1,\dots,f_{n-N_0} $, which is $\im \psi^N$. Thus, $\phi: \im \psi^N \to \im \psi^N$ is surjective, and the result follows.
\end{proof}

\begin{rem}[how to do it in multigraded case]
  Let $R$ be an $\ZZ^n$-graded ring, with $R_0=k$ an algebraically closed field.
  Assume furthermore that the degree support of $R$ is strongly convex, i.e., that
  the effective cone
  $\Eff R = \{ \deg r : r\in R\}$ is a strongly convex cone in $\ZZ^n$.
  Let $M$ be a finitely generated graded $R$-module, and let $\phi \in \End M$ be a degree zero endomorphism.

  Let $P\subset \ZZ^n$ be the degrees of a set of minimal generators of $M$;
  this is a poset with respect to the partial order $\a\leq \b$ if and only if $\b-\a\in \Eff R$.
  Because $P$ is finite, we can order the elements of $P$ such that
  $\deg m_1\not\leq \deg m_2\not \leq \dots\not\leq \deg m_r$.
  \mahrud{finish}
\end{rem}

The above can of course produce the trivial direct sum decomposition: if the eigenvalues of $\phi$ are all nonzero, the proof yields that $\im \psi = M$ and $\coker \psi = 0$. The following corollary of the proof of \cref{lem:graded_algorithm} ensures that we can always assume $\phi$ has zero as an eigenvalue:

\begin{lem}
  The eigenvalues of $\phi$ lie in $k$. In particular, for any $\phi \in [\End M]_0$, there is $\lambda \in k$ such that $\phi - \lambda \id_M \in [\End M]_0$ has zero as an eigenvalue.
\end{lem}

\mahrud{move this above?}
By ``eigenvalues of $\phi$'' we mean the roots of the characteristic polynomial $\det(\phi - \lambda \id_M)$\mahrud{define determinant over a module, or say determinant of its cover?}, which is a univariate polynomial in $\lambda$ with coefficients in $R$ and thus a priori has solutions only in the algebraic closure of the total ring of fractions of $R$.

\begin{proof}
  The expression of $\phi$ obtained in \cref{eq:Jordan} shows that the determinant of $\phi - \lambda \id_M$ is the product of $\lambda_i - \lambda$, thus the eigenvalues are the $\lambda_i$, which all lie in $k$.
\end{proof}

@@

\begin{exa}[The local algorithm]
  Let $R=\F_2[x,y,z]_{(x,y,z)}/(x^2y+xy^2+xyz+z^2)$. In the notation of \cite{Artin77}, this is the $D_4^1$ singularity, which is an exceptional version of the usual rational double point/Du Val singularities appearing in characteristic 2.  In particular, note that $R$ is not homogeneous. If $F_* R$ denotes the Frobenius pushforward of $R$, then the algorithm of \Cref{sec:local-alg} computes the following indecomposable summands:
  $$
  F_*R = R\oplus
  \coker
  \begin{pmatrix}
    x+y+z&z\\
    z&xy
  \end{pmatrix}\oplus\coker \begin{pmatrix}
    y&z\\
    z&x^{2}+xy+xz
  \end{pmatrix}\oplus \coker\begin{pmatrix}
  x&z\\
  z&xy+y^{2}+yz
  \end{pmatrix};
  $$
  there is one free summand and three reflexive modules of rank 1.
\end{exa}

>>>>

\begin{lem}\label{lem:graded}
  Let $N = \mu(M)$ and set $\psi = \phi^N$. Then $M = \im\psi \oplus \coker\psi$.
\end{lem}
\begin{proof}
  Fix homogeneous generators $m_1,\dots,m_N$ of $M$, with $\deg m_i$ nonincreasing, and let $d_1,\dots,d_n$ be the decreasing list of unique degrees of generators of $M$. The matrix for $\phi$ is then ``block upper triangular'' with $n$ diagonal blocks with entries in $k$, i.e., of the form
  \[ \phi = \begin{pmatrix}
    B_1    & *      & \dots  & *      \\
    0      & B_2    & \dots  & *      \\
    \vdots & \vdots & \ddots & \vdots \\
    0      & 0      & \dots  & B_n
  \end{pmatrix}, \]
  where the blocks $B_i$ are square matrices \emph{with entries in $k$}, and the remaining entries are in $R$. To see this, note that since $\phi$ has degree zero, $\phi(m_i)$ is a $k$-linear combination of the $m_j$ with $\deg m_j = \deg m_i$ plus an $R$-linear combination of the $m_j$ with $\deg m_j < \deg m_i$. In particular, $B_i$ is the induced endomorphism on the submodule generated by all generators of degree $d_i$.

@@

  Let $\alpha$ be the composition of these automorphisms for each block $B_i$, then $\phi' = \alpha^{-1}\circ\phi\circ\alpha$ corresponds to performing row and column operations on $\phi$ over $k$ and has the form
  \begin{equation}
    \label{eq:Jordan}
    \phi' = \begin{pmatrix}
      J_1    & *      & \dots  & *      \\
      0      & J_2    & \dots  & *      \\
      \vdots & \vdots & \ddots & \vdots \\
      0      & 0      & \dots  & J_n
    \end{pmatrix},
  \end{equation}
  where each $J_i$ is a Jordan block over $k$ with eigenvalue $\lambda_i$, and the remaining entries are in $R$.
  Let $\mu_i$ be the multiplicity of $\lambda_i$ and let $\psi_i = (\phi' - \lambda_i \id_M)^{\mu_i}$.
  \mahrud{TODO}

  %% Relabeling, we may assume that $\lambda_i\neq 0$ for $i=1,\dots,n_0$ and $\lambda_i=0$ for $i=n_0+1,\dots,n$.

  %% If we take $N_0$ to be at least the sum of the sizes of the nilpotent Jordan blocks $J_{n_0+1},\dots,J_n$, then for $N\geq N_0$ we have
  %% $$
  %% \phi^N =
  %% \begin{pmatrix}
  %%   J_1^N & *** &\dots & \dots& \dots  &\dots & *** \\
  %%   0 & J_2^N & \dots &\dots&\dots & \dots& *** \\
  %%   \vdots &\vdots & \ddots &\vdots \\
  %%   0 & 0 & \dots & J_{n_0}^N & *** & \dots & ***\\
  %%   0 & 0 & \dots & 0 & 0 & \dots & 0\\
  %%   \vdots &\vdots & \ddots &\vdots \\
  %%   0 & 0 & \dots & 0 & 0 & \dots & 0
  %% \end{pmatrix},
  %% $$
  %% where each nonzero $J_i^N$ has the units $\lambda_i^N$ on the diagonal.
  %% ($\mu(M)$ is an upper bound on the sum of the sizes of the nilpotent Jordan blocks, so the above holds for $N\geq \mu(M)$.)

  %% Fix $N$ as above.
  %% Now, consider the map $ \phi^N: \im(\phi^N) \to \im(\phi^N)$. If this map is surjective, then $\im(\phi^N)$ is a direct summand of $M$, with complementary summand $\coker \phi^N$: to see this, observe that the injection $\im(\phi^N) \inc M$ would then  compose with $\phi^N$ to be a surjection, and since $\im(\phi^N)$ is a finitely generated module, a surjection must be an isomorphism $\a:\im \phi^N \to \im \phi^N$ by \cite[Corollary~4.4]{Eisenbud95}. Thus, we obtain a splitting of the inclusion $\im \phi^N \inc M$ via $\a\inv\circ \phi^N$; since $\a\inv$ is an isomorphism, we have that $\im \phi^N$ is (isomorphic to) a direct summand of $M$, with complementary summand $\coker \phi^N$.

  %% So, all that needs to be seen is that $\psi:=\psi^N$ is surjective.
  %% To see this, note that after replacing the original generators for $M$ by the generators that put $\psi$ into the form above,  a generating set for $\im(\phi^N)$ is given by $f_i = \psi(m_i) = u_i\cdot e_i + \sum_{j<i } a_{ij} e_j$ for $i = 1,\dots,n-N_0$, where $u_i$ is a unit (explicitly, $u_i$ is a power of some nonzero $\lambda_{i'}\in k$).
  %% That is, a generating set is given by
  %% $$u_1 e_1,\quad  u_2 e_2+a_{21}e_1,\quad \dots ,\quad u_{n-n_0}e_{n-N_0}+a_{n-N_0,n-1}e_{n-N_0-1}+\dots + a_{N-n_0,1}e_1.$$
  %% Since the $u_i$ are units, the $R$-span of these generators is clearly the same as the $R$-span of $e_1,\dots,e_{n-N_0}$, and thus after applying $\psi$ we get back the $R$-span of $f_1,\dots,f_{n-N_0} $, which is $\im \psi^N$. Thus, $\phi: \im \psi^N \to \im \psi^N$ is surjective, and the result follows.
\end{proof}

\begin{rem}
  The hypotheses that $R$ is $\N$-graded and $k$ is an algebraically closed field may both be weakened.
  In the proof of \Cref{lem:graded}, we only need a partial order on the degrees of the generators of $M$ in order to present $\phi$ in a block upper triangular form. Suppose $R$ is $\ZZ^r$-graded, then the effective cone $\Eff(R)\subset\Q^n$ generated by the elements $\deg m$ for each monomial $m \in R$ induces a partial order on the degrees of generators of $M$ if it is strongly convex (contains no positive dimensional subspace of $\Q^n$) and $R_0 = k$. % TODO: can R_0 be finite dimensional?

  Similarly, suppose $k$ is not algebraically closed, then there may be blocks $B_i$ which cannot be written in Jordan canonical form. Nevertheless, as long as $\phi$ has at least one eigenvalue $\lambda$ in $k$, the corresponding block may be written in Jordan canonical form, and thus the image and cokernel of $(\phi - \lambda_i\id_M)^{\mu_i}$ form summands of $M$.
\end{rem}

% The above can of course produce the trivial direct sum decomposition: if the eigenvalues of $\phi$ are all nonzero, the proof yields that $\im \psi = M$ and $\coker \psi = 0$. The following corollary of the proof of \Cref{lem:graded} ensures that we can always assume $\phi$ has zero as an eigenvalue:

%% \begin{lem}
%%   The eigenvalues of $\phi$ lie in $k$. In particular, for any $\phi \in [\End M]_0$, there is $\lambda \in k$ such that $\phi - \lambda \id_M \in [\End M]_0$ has zero as an eigenvalue.
%% \end{lem}

By ``eigenvalues of $\phi$'' we mean the roots of the characteristic polynomial $\det(\phi - \lambda \id_M)$\mahrud{define determinant over a module, or say determinant of its cover?}, which is a univariate polynomial in $\lambda$ with coefficients in $R$ and thus a priori has solutions only in the algebraic closure of the total ring of fractions of $R$.

%% \begin{proof}
%%   The expression of $\phi$ obtained in \cref{eq:Jordan} shows that the determinant of $\phi - \lambda \id_M$ is the product of $\lambda_i - \lambda$, thus the eigenvalues are the $\lambda_i$, which all lie in $k$.
%% \end{proof}

@@

\begin{exa}[Local singularities]
  Let $R=\F_2[x,y,z]_{(x,y,z)}/(x^2y+xy^2+xyz+z^2)$. In the notation of \cite{Artin77}, this is the $D_4^1$ singularity, which is an exceptional version of the usual rational double point/Du Val singularities appearing in characteristic 2.  In particular, note that $R$ is not homogeneous. If $F_* R$ denotes the Frobenius pushforward of $R$, then the algorithm of \Cref{sec:local-alg} computes the following indecomposable summands:
  $$
  F_*R = R\oplus
  \coker
  \begin{pmatrix}
    x+y+z&z\\
    z&xy
  \end{pmatrix}\oplus\coker \begin{pmatrix}
    y&z\\
    z&x^{2}+xy+xz
  \end{pmatrix}\oplus \coker\begin{pmatrix}
  x&z\\
  z&xy+y^{2}+yz
  \end{pmatrix};
  $$
  there is one free summand and three reflexive modules of rank 1.
\end{exa}
